---
title: "Cell cycle genes"
author: "Matthew Stephens"
date: "April 9, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Some Cell Cycle Data

The data come from a recent experiment performed in the Gilad lab by Po Tung, in collaboration with Joyce Hsiao and others. 

Since this is an anlaysis in progress, there is a lot we do not
know about these data.


The data are measuring the activity of 10 genes that may or may not be involved in the "cell cycle", which is the process cells go through as
they divide. (We have data on a large number of genes, but Joyce has picked out 10 of them for us to look at.) Each gene is measured
in many single cells, and we have some independent (but noisy) measurement of where each cell is in the cell cycle.


```{r}
d = readRDS("../data/cyclegenes.rds")
dim(d)
```

Here each row is a single cell. The first column ("theta") is an estimate
of where that cell is in the cell cycle, from 0 to 2pi. (Note
that we don't know what stage of the cell cycle each point in the
interval corresponds to - so there is no guarantee that 0 is the
"start" of the cell cycle. Also, because of the way these data were created we don't know which direction the cell cycle is going - it could be forward or backward.) Then there are 10 columns corresponding
to 10 different genes.

Here we just plot 8 genes to get a sense for the data:
```{r}
par(mfcol=c(2,2))
for(i in 1:4){
  plot(d$theta, d[,(i+1)],pch=".",ylab="gene expression",xlab="cell cycle location")
}
par(mfcol=c(2,2))
for(i in 1:4){
  plot(d$theta, d[,(i+5)],pch=".",ylab="gene expression",xlab="cell cycle location")
}
```

The question we want to answer is this: which genes show greatest
evidence for varying in their expression through the cell cycle?
For now what we really want is a filter we can apply to a large
number of genes to pick out the ones that look most "interesting".
Later we might want a more formal statistical measure of evidence.


To apply wavelets we need the data to be a power of 2.
Also we need the data to be ordered in terms of theta.
We'll subset the data to 512 elements here, and order it:
```{r}
# order the data
o = order(d[,1])
d = d[o,]
plot(d$theta)

# subset the data
set.seed(1)
subset = sort(sample(1:nrow(d),512,replace=FALSE))
d.sub = d[subset,]
```


Here we do the Haar wavelet by specifying `family="DaubExPhase",filter.number = 1` to the discrete wavelet 
transform function `wd`. The plot shows the wavelet transformed
values, separately at each resolution.
```{r}
library("wavethresh")
wds <- wd(d.sub[,2],family="DaubExPhase",filter.number = 1)
plot(wds)
```

To illustrate the idea behind wavelet shrinkage we use
the policy "manual" to shrink all the high-resolution coefficients 
(levels 4-8) to 0. 
```{r}
wtd <- threshold(wds, levels = 4:8,  policy="manual",value = 99999) 
plot(wtd) 
```

Now undo the wavelet transform on the shrunken coefficients
```{r}
fd <- wr(wtd) #reconstruct
plot(d.sub$theta,d.sub[,2],xlab="cell cycle", ylab = "Expression")
lines(d.sub$theta,fd,col=2,lwd=3)
```


The estimate here is a bit "jumpy", due to the use of the Haar wavelet and the rather naive hard thresholding. We can make it less "jumpy"
by using a "less step-wise" wavelet basis
```{r}
wds <- wd(d.sub[,2],family="DaubLeAsymm",filter.number = 8)
wtd <- threshold(wds, levels = 4:8,  policy="manual",value = 99999) 
fd <- wr(wtd) #reconstruct
plot(d.sub$theta,d.sub[,2],xlab="cell cycle", ylab = "Expression")
lines(d.sub$theta,fd,col=2,lwd=3)
```

What happens if we use an Empirical Bayes thresholding rule?

The EbayesThresh package essentially solves the EBNM problem,
using a prior distribution that is a mixture of a point mass at 0
and a Laplace distribution with rate parameter $a$:
$$ \pi_0 \delta_0 + (1-\pi_0) DExp(a).$$

Here `a=NA` tells Ebayesthresh ot estimate $a$.
Notice that the outliers cause "problems"
```{r}
library("EbayesThresh")
wds <- wd(d.sub[,2],family="DaubLeAsymm",filter.number = 8)
wtd <- ebayesthresh.wavelet(wds,a=NA)
fd <- wr(wtd) #reconstruct
plot(d.sub$theta,d.sub[,2],xlab="cell cycle", ylab = "Expression")
lines(d.sub$theta,fd,col=2,lwd=3)
```

Try removing the outliers (actually setting them to the mean here),
just to see what happens.
```{r}
xx = ifelse(d.sub[,2]<2,mean(d.sub[,2]),d.sub[,2])
wds <- wd(xx,family="DaubLeAsymm",filter.number = 8)
wtd <- ebayesthresh.wavelet(wds,a=NA)
fd <- wr(wtd) #reconstruct
plot(d.sub$theta,xx,xlab="cell cycle", ylab = "Expression")
lines(d.sub$theta,fd,col=2,lwd=3)
```


Notice that estimating $a$ is really important (the default is
to set $a=0.5$):
```{r}
xx = ifelse(d.sub[,2]<2,mean(d.sub[,2]),d.sub[,2])
wds <- wd(xx,family="DaubLeAsymm",filter.number = 8)
wtd <- ebayesthresh.wavelet(wds)
fd <- wr(wtd) #reconstruct
plot(d.sub$theta,xx,xlab="cell cycle", ylab = "Expression")
lines(d.sub$theta,fd,col=2,lwd=3)
```




