---
title: "currency analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this document I read in some currency data, provide a brief explanation
of the data, and illustrate some preliminary transformations one might perform. I had the intention of using these for a homework but my initial 
analyses suggested the structure in the data are a bit subtle so I decided not
to use them in homework for now.

## Read in data

The data were downloaded and saved in the `data` directory as
documented [here](currency.html). They consist of daily exchange rates
(vs the US dollar, abbreviated USD) for 191 currencies. Note
that the first column is USD itself, so exchange rate is always 1.
Note also that the first row is the most recent day, and last row is about 6 months ago; so that the vector of observations is "forwards in time" I reverse the rows here.

```{r}
df = read.csv("../data/currency.csv")
df = apply(df,2,rev)
head(df)
```

## Abbreviations

You can find the meanings for the abbreviations using the `quantmod` library:
```{r}
currency_table = quantmod::oanda.currencies
currency_table
```


## Processing

We are going to use these data to look at covariances of the
change in exchange rates. Of course, for a given currency,
the exchange rate from one day
to the next tend to be similar -- that is the time series will show strong
autocorrelation. Here's an example time series for the Afghanistan Afghani (AFN):
```{r}
plot(df[,"AFN"],type="l", main="Exchange rates for AFN vs USD",ylab="day")
```

On the other hand, one might expect the *changes* in exchange rate from day to day to be less correlated. So I suggest analysing the changes from day to day.
Also these data are positive, and it makes sense to measure changes on
a multiplicative scale, so I will take the log and the look at the differences.

```{r}
ldf = log(df)
ldf_diff = apply(ldf,2,diff)
plot(ldf_diff[,"AFN"], type="l", main = "change in log exchange rate by day, AFN vs USD")
```

And we can look at how correlated these are between currencies. Eg when the 
AFN goes up vs the dollar, does the Albanian Lek (ALL) also go up vs dollar?
Maybe there is a hint of a positive correlation here? (One might
expect a correlation as they are both being measured against the same thing, the USD... if something happens to weaken the USD then maybe both would go up....)
```{r}
plot(ldf_diff[,"AFN"], ldf_diff[,"ALL"])
cor(ldf_diff[,"AFN"], ldf_diff[,"ALL"])
```


