---
title: "histogram for cell cycle"
author: "Matthew Stephens"
date: "April 11, 2018"
output: workflowr::wflow_html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

We will try estimating the density of the cell cycle values.
See (cell_cycle.Rmd) for more on those data.

# Histograms

First we try histograms. The histograms here are un-normalized, so not actually densities. We could normalize them
easily enough, but I think it is helpful for you to see the counts get smaller.

Let's start with the default:
```{r}
d = readRDS("../data/cyclegenes.rds")
hist(d[,1],xlab="theta (cell cycle position)", main="Histogram of theta")
```

And now try decreasing bin sizes to see the bias/variance trade-off in action.
```{r}
hist(d[,1],breaks = seq(0,2*pi,length=5),xlab="theta (cell cycle position)", main="Histogram of theta")
hist(d[,1],breaks = seq(0,2*pi,length=20),xlab="theta (cell cycle position)", main="Histogram of theta")
hist(d[,1],breaks = seq(0,2*pi,length=100),xlab="theta (cell cycle position)", main="Histogram of theta")
hist(d[,1],breaks = seq(0,2*pi,length=400),xlab="theta (cell cycle position)", main="Histogram of theta")
hist(d[,1],breaks = seq(0,2*pi,length=1000),xlab="theta (cell cycle position)", main="Histogram of theta")
```


# Density estimation

Let's try the default density estimation in the R function "density".
```{r}
hist(d[,1],breaks = seq(0,2*pi,length=100),xlab="theta (cell cycle position)", main="Histogram of theta",probability=TRUE)
d.dens = density(d[,1])
lines(d.dens$x,d.dens$y,col=2,lwd=2)
```

The analogue of bin size is "bandwidth", set using bw.
Here we try four very different bandwidths.
```{r}
hist(d[,1],breaks = seq(0,2*pi,length=100),xlab="theta (cell cycle position)", main="Histogram of theta",probability=TRUE)
d1.dens = density(d[,1],bw=0.01)
d2.dens = density(d[,1],bw=0.1)
d3.dens = density(d[,1],bw=1)
d4.dens = density(d[,1],bw=10)

lines(d1.dens$x,d1.dens$y,col=2,lwd=2)
lines(d2.dens$x,d2.dens$y,col=3,lwd=2)
lines(d3.dens$x,d3.dens$y,col=4,lwd=2)
lines(d4.dens$x,d4.dens$y,col=5,lwd=2)
```

# A common issue with density

Go back to the origin plot. 
```{r}
hist(d[,1],breaks = seq(0,2*pi,length=100),xlab="theta (cell cycle position)", main="Histogram of theta",probability=TRUE)
d.dens = density(d[,1])
lines(d.dens$x,d.dens$y,col=2,lwd=2)
```


Notice how the density seems too low at both
ends of the plot. This is a common problem with density when applied to a finite domain like this. It is because it is estimating the density at 0 and notices
there are no observations just to the left of 0, and this biases the estimate of density down.

We can see the problem better if we plot a wider area:
```{r}
hist(d[,1],breaks = seq(-pi,3*pi,length=100),xlab="theta (cell cycle position)", main="Histogram of theta",probability=TRUE)
d.dens = density(d[,1])
lines(d.dens$x,d.dens$y,col=2,lwd=2)
```


To deal with this a simple trick is to reflect the data at both ends.

```{r}
x = d[,1]
x = c(-x,x,2*pi + 2*pi-x)
hist(x,prob=TRUE)
```

er... yuk! the default histogram was not very nice - have to do it by hand!
```{r}
hist(x,breaks = seq(-2*pi, 4*pi,length=100),probability=TRUE)
abline(v=0,col=2)
abline(v=2*pi,col=2)
```

So now we can estimate the density
```{r}
x.dens = density(x)
hist(x,breaks = seq(-2*pi, 4*pi,length=100),probability=TRUE)
abline(v=0,col=2)
abline(v=2*pi,col=2)
lines(x.dens$x,x.dens$y,col=2)
```

Also yuk!! I had to experiment with the parameter `bw` to get a nicer plot....
```{r}
x.dens = density(x,bw=0.2)
hist(x,breaks = seq(-2*pi, 4*pi,length=100),probability=TRUE)
abline(v=0,col=2)
abline(v=2*pi,col=2)
lines(x.dens$x,x.dens$y,col=3)
```

I now could get a nice density estimate in the range 0 to 2pi by just truncating and renormalizing to have area 1.



