---
title: "GP_cell_cycle"
author: "Matthew Stephens"
date: "5/21/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I'm going to try fitting a GP to the cell cycle data:

```{r}
d = readRDS("../data/cyclegenes.rds")
dim(d)
```

Recall each row is a single cell. The first column ("theta") is an estimate
of where that cell is in the cell cycle, from 0 to 2pi. (Note
that we don't know what stage of the cell cycle each point in the
interval corresponds to - so there is no guarantee that 0 is the
"start" of the cell cycle. Also, because of the way these data were created we don't know which direction the cell cycle is going - it could be forward or backward.) Then there are 10 columns corresponding
to 10 different genes.

I'm going to order the rows by cell cycle (theta, first column) as this will make
things much easier later.
```{r}
# order the data
o = order(d[,1])
d = d[o,]
plot(d$theta)
```



# Trend filtering

Remember before we tried trend filtering on these data...

```{r}
library(genlasso)
d2.tf = trendfilter(d[,2],ord = 1)
d2.tf.cv = cv.trendfilter(d2.tf) # performs 5-fold CV
plot(d[,1],d[,2],xlab="cell cycle",ylab="expression")
lines(d[,1],predict(d2.tf, d2.tf.cv$lambda.min)$fit,col=2,lwd=3)
```

I found GP_fit took a while, so I subsampled to 200 to try to reduce time. Also it seemed to want $x$ in the range $[0,1]$ so I normalized the $x$ values for it.
```{r}
subset = sort(sample(1:nrow(d),200))
x = d[subset,1]
y = d[subset,2]
x = x/max(x) # normalize to be in 0,1
y.gpfit = GPfit::GP_fit(x,y)
plot(x,y)
lines(x,predict(y.gpfit)$Y_hat,col=2)
```

