---
title: "Gaussian process"
author: "Matthew Stephens"
date: "5/21/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulate Gaussian Process

Here we simulate a GP with squared exponential kernel:

```{r}
set.seed(1)
x = seq(0,1,length=100)
d = abs(outer(x,x,"-")) # compute distance matrix, d_{ij} = |x_i - x_j|
l = 1 # length scale
Sigma_SE = exp(-d^2/(2*l^2)) # squared exponential kernel
y = mvtnorm::rmvnorm(1,sigma=Sigma_SE)
plot(x,y)
```

Try making the covariance decay faster with distance:
```{r}
l = 0.1
Sigma_SE = exp(-d^2/(2*l^2)) # squared exponential kernel
y = mvtnorm::rmvnorm(1,sigma=Sigma_SE)
plot(x,y)
```

Here is a plot of five different simulations:
```{r}
plot(x,y,type="l",ylim=c(-3,3))
for(i in 1:4){
  y = mvtnorm::rmvnorm(1,sigma=Sigma_SE)
  lines(x,y,col=i+1)
}
```

# The OU covariance function:

Here we use the covariance function for 
what is known as the "Ornstein--Uhlenbeck process",
which you can think of as a modified Brownian motion,
where the modification tends to pull the process back towards 0.
(Unmodified BM tends to wander progressively further from 0.)

Notice
it produces much "rougher" functions (actually not differentiable)!
```{r}
Sigma_OU = exp(-d/l) # OU kernel
y = mvtnorm::rmvnorm(1,sigma=Sigma_OU)
plot(x,y,type="l",ylim=c(-3,3))
for(i in 1:4){
  y = mvtnorm::rmvnorm(1,sigma=Sigma_OU)
  lines(x,y,col=i+1)
}
```


# Matern covariance function

```{r}
library("geoR")
Sigma_M = matern(d,phi=l,kappa=1) 
y = mvtnorm::rmvnorm(1,sigma=Sigma_M)
plot(x,y,type="l",ylim=c(-3,3))
for(i in 1:4){
  y = mvtnorm::rmvnorm(1,sigma=Sigma_M)
  lines(x,y,col=i+1)
}
```

