---
title: "stocks"
author: "Matthew Stephens"
date: "2021-05-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Here I download and save some stock price data. I got some help from
https://www.codingfinance.com/post/2018-03-27-download-price/

Here are the stocks I download:
```{r}
# AAPL: Apple
# NFLX: Netflix
# AMZN: Amazon
# MMM: 3M
# K: Kellogs
# O: Realty Income Corp
# NSRGY: Nestle
# LDSVF: Lindt
# JPM: JP Morgan Chase
# JNJ: Johnson and Johnson
# TSLA: Tesla
# V: Visa
# PFE: Pfizer
```

## Downloading and Saving

Here I use the quantmod package to download and save the data:
```{r}
library(tidyquant)
library(purrr)


tickers = c("AAPL", "NFLX", "AMZN", "MMM", "K", "O", "NSRGY", "LDSVF", "JPM", "JNJ", "TSLA", "V", "PFE")
getSymbols(tickers)

prices <- map(tickers,function(x) Ad(get(x))) # gets the adjusted prices of each stock
prices <- reduce(prices,merge)
colnames(prices) <- tickers
head(prices)
```

Some companies (eg TLSA) were not listed for the entire period avoilable.
I'm going to narrow down the time window so no missing data.
This ends up with the period 2010-10-15 to 2021-05-21.
```{r}
nomiss = function(x){all(!is.na(x))}
prices = prices[apply(prices,1,nomiss),]
head(prices)
tail(prices)
write.csv(prices, file="../data/prices.csv",quote=FALSE,row.names=FALSE)
```


## Reading in and processing

Here are brief suggestions for reading and processing the data.
First since stocks are positive, and
what generally matters in stocks is percentage change in price,
it makes sense to take logs and look at differences.
These are called "log-returns" in finance.
Here I  load in the data and compute the log returns.

```{r}
prices = read.csv("../data/prices.csv")
log_prices = log(prices)
log_returns = apply(log_prices,2, diff)
```


Note that while the stock prices from day to day are highly correlated
(as they form a time series), log returns are much less correlated
(and so easier to study and model statistically).

```{r}
plot(log_prices[,1], main="Apple (log) stock price over time")
plot(log_returns[,1], main="Apple log-returns over time")
```

One question of interest is to what extent
changes in stocks are correlated with one another.
Here is an initial plot showing some of the correlation
structure in the data....
```{r}
S = cor(log_returns)
heatmap(S, xlab = names(prices), symm=TRUE)
```


