---
title: "sudden_changes"
author: "Matthew Stephens"
date: "2022-04-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(smooth)
library(genlasso)
library(smashr)
```

## Introduction

I wanted to illustrate the idea that some smoothing techniques are better than others
at dealing with changes that occur on different "scales" (eg a combination of smooth changes as sudden changes).

## Simulated data

Here I set up $E(y)=\mu$ as a function that varies smoothly except for a sudden change in the middle, and simulate some data with that mean:

```{r}
t = seq(0,1,length=1024)
mu = (t-0.5)^2 - 0.2*(t>0.5)

y = rnorm(1024,mu,0.05)
plot(t,mu, type="l", col=2, lwd=2)
points(t,y)
```


## Compare some smoothers

## Simple moving average (sma)

I tried a sliding window of sizes 10 and 50:
```{r}
y.sma.10 = sma(y,order=10) 
y.sma.50 = sma(y,order=50)
plot(t,mu, type="l", col=2, lwd=2, main="sma, order 10 (green) and 50 (blue)")

lines(t,y.sma.10$fitted,col=3,lwd=2)
lines(t,y.sma.50$fitted,col=4,lwd=2)
```

## Trend-filtering
```{r}
y.tf.0 = trendfilter(y,t,ord=0)
y.tf.0.cv = cv.trendfilter(y.tf.0) # performs 5-fold CV

y.tf.1 = trendfilter(y,t,ord=1)
y.tf.1.cv = cv.trendfilter(y.tf.1) # performs 5-fold CV


plot(t,mu, type="l", col=2, lwd=2, main="trendfilter (order 0 = green; order 1=blue)")

lines(t,predict(y.tf.0, y.tf.0.cv$lambda.min)$fit,col=3,lwd=2)
lines(t,predict(y.tf.1, y.tf.1.cv$lambda.min)$fit,col=4,lwd=2)
```

## Wavelet with EB shrinkage

```{r}
y.smash = smash(y,filter.number=1) 
plot(t,mu, type="l", col=2, lwd=2, main="Wavelet (haar)")
lines(t,y.smash,col=3,lwd=2)
```

