---
title: "svd_single_cell_data"
author: "Matthew Stephens"
date: "2021-05-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

I wanted to illutrate SVD on the single cell data we used in the 2021
midterm.

Read in the data:
```{r}
df = read.csv("../data/cell_data.csv",sep=",")
X = as.matrix(df[,-1])
head(df)
```

Here I simply run svd and plot the first two eigenvectors against one another.
(One could imagine wanting to standardize the columsn first, but I did not do that for now).
```{r}
X.svd = svd(X)
plot(X.svd$u[,1],X.svd$u[,2],col=c(rep(1,1000),rep(2,1000)), main = "first PC vs second PC, colored by cell type")
```

So it is seems that the cell types
separate somewhat in PC space (even though there is considerable overlap). And yet the
EM algorithm for Poisson clustering did not result in these two clusters. Why not?

Have a look at PC1. It turns out that this is basically the
total counts for each cell:
```{r}
plot(X.svd$u[,1],rowSums(X))
```

So what is going on is this. the Poisson cluster model
ends up splitting the cells on PC1, basically splitting the cells
by the total counts. This is like splitting documents into groups based
on how long they are (how many words) rather than what the words say...
PC2 is more about what the words say.

Basically, if we want to model based approach to have a chance to 
split into two groups we need to *control for the fact that different cells
have different total counts*
in the model.





