---
title: "Run svd on digits data"
author: "Matthew Stephens"
date: "April 17, 2018"
output: html_document
---

## Introduction

Here I run both SVD and non-negative matrix factorization on the zipcode training data.

## SVD


Here we read in the "zipcode" training data, and extract the 2s and 3s.
```{r}
z = read.table("../data/zip.train.txt")
sub = (z[,1] == 2) | (z[,1]==3)
z23 = as.matrix(z[sub,])
```

Now we run svd (excluding the first column which are the labels)
```{r}
z23.svd = svd(z23[,-1])
```


Plot the first two two singular vectors, colored by group,
we see the second sv separates the groups reasonably well.
```{r}
plot(z23.svd$u[,1],z23.svd$u[,2],col=z23[,1])
```

And a histogram suggests a mixture of two Gaussians might be 
a reasonable start:
```{r}
hist(z23.svd$u[,2],breaks=seq(-0.07,0.07,length=20))
```

# Plot the images corresponding to first two eigenvectors

We put the eigenvector in the form of an image before plotting.
Note that I have not been very careful with the limits
```{r}
n = sqrt(length(z23.svd$v[,1]))
plot_zipcode_image = function(v){
  n = sqrt(length(v))
  m = matrix(v,nrow=n)
  m = apply(m,1,rev) # I had to experiment here to get the images to appear the "right way up"
  m = apply(m,1,rev) # note that applying rev twice does not result in no change...
  m = apply(m,2,rev)
  image(m)
}
plot_zipcode_image(z23.svd$v[,1])
plot_zipcode_image(z23.svd$v[,2])
```

# Try non-negative matrix factorization

You will need to install `NNLM` to run this.
```{r}
#library(devtools)
#devtools::install_github('linxihui/NNLM')

library(NNLM)
summary(as.vector(z23[1,]))
z23_nonneg = z23[,-1] + 1
z23.nmf = nnmf(z23_nonneg,k = 2)
plot_zipcode_image(z23.nmf$H[1,])
plot_zipcode_image(z23.nmf$H[2,])
plot(z23.nmf$W[,1],z23.nmf$W[,2],col=z23[,1])
```


Try doing 6 and 3 instead - you see it separates them very nicely.
```{r}
sub = (z[,1] == 6) | (z[,1]==3)
z36 = as.matrix(z[sub,])
summary(as.vector(z36[1,]))
z36_nonneg = z36[,-1] + 1
z36.nmf = nnmf(z36_nonneg,k = 2)
plot_zipcode_image(z36.nmf$H[1,])
plot_zipcode_image(z36.nmf$H[2,])
plot(z36.nmf$W[,1],z36.nmf$W[,2],col=z36[,1])
```


Try doing all the digits. Here the results are messier. It is possible
this is a local optimum problem as nmf is sensitive to initialization.
```{r}
z_nonneg = as.matrix(z[,-1]) + 1
z.nmf = nnmf(z_nonneg,k = 12,max.iter = 100)
par(mfcol=c(3,4),mai=rep(0.2,4))
for(i in 1:12){
  plot_zipcode_image(z.nmf$H[i,])
}
```


Try running multiple times. Here I plot the mse trace for 3 different runs.

```{r}
z.nmf2 = nnmf(z_nonneg,k = 12,max.iter = 100)
z.nmf3 = nnmf(z_nonneg,k = 12,max.iter = 100)
par(mfcol=c(1,1),mai=rep(0.5,4))
plot(z.nmf$mse,type="l")
lines(z.nmf2$mse,col=2)
lines(z.nmf3$mse,col=3)

```


